Accelerating adaptive ultrasound algorithms using graphics processing units \vspace{0.5cm}

### 2014-03-08 ThesisJP/Background/background.tex
s

Modern GPUs can also be interpreted as a unit consisting of multiple SIMD processors (Fig.\,\ref{fig:cpu_gpu}). The Kepler architecture by Nvidia have up to 15 SIMD units capable of processing 192, 32 bit, floats per clock cycle at 745 MHz. These GPUs are also capable of performing FMA instructions, effectively doubling the number of theoretical FLOPS. The theoretical peak single precision FLOPS for this architecture is by that found to be 
\begin{equation}
15*192*2*745 = 4291\,\text{GFLOPS}.
\end{equation} 
Each SIMD core also comprise 16 special function units (\nom{SFU}{Special function unit (sin, exp, sqrt, ect.)}) which add close to 1 TFLOPS in additional computational power. This adds up to a total of five TFLOPS as shown in Fig.\,\ref{fig:cpu_vs_gpu} for the GeForce GTX 780 TI GPU. From Fig.\,\ref{fig:cpu_gpu} we see how the high number of SIMD cores found on GPUs and their larger width, compared with the SIMD cores on CPUs, results in five times the computational performance of a high-end CPU (without the integrated GPU).

@@

Both CUDA and OpenCL is centered around a kernel function which are executed by a large group of threads. The kernel function, for both CUDA and OpenCL, has a C-like syntax with some additional specifiers. A CUDA kernel is depicted in Fig.\,\ref{fig:gpu_kernel}. The \texttt{\_\_global\_\_} specifier tells the CUDA compiler that this function is a kernel function, and in CUDA, this kernel function is launched by using a CUDA specific syntax:
\begin{align*}
\texttt{square<<<grid, block>>>(a, N)},
\end{align*}
where \texttt{grid} and \texttt{block} is 3-component vectors specifying grid and block size in "xyz" as depicted in Fig.\,\ref{fig:gpu_grid}, \texttt{a} is a pointer to a memory buffer located on the GPU, and \texttt{N}$*$\texttt{N} is the buffer size. When launched, a copy of the kernel is scheduled  for each thread in each block, and each kernel gets as input its location in the compute grid (\texttt{blockIdx} and \texttt{threadIdx}) together with the grid size (\texttt{gridDim} and \texttt{blockDim}). This information is then typically used to calculate which memory elements to gather from global memory, in this case the elements in \texttt{a}. %Behind the scene, thread blocks are divided into smaller sets of threads called a warp (or wavefront in OpenCL). 
When a kernel is executed, multiple thread blocks are scheduled to each SIMD core where they stay until all instructions in the kernel is finished for all threads in the block. Thus,  in order to formulate a given problem as a kernel function the global memory input to each thread must not rely on the output from other blocks, hence it must be possible for the GPU to scheduled blocks in any order. As we see, a thread has a different conceptual meaning on a GPU than on a CPU.  Where a CPU is said to execute threads which can perform SIMD instructions, a thread on a GPU can be interpreted as stream capable of processing one data element in the SIMD instruction. This is why GPU threads are said to be light weighted.

The code in Fig.\,\ref{fig:gpu_grid} is of course a minimal example of GPU programing. A full example will involve several lines of code transferring data to and from the GPU, and the kernel could include manual administration of the L1 cache (shared memory), and intra-block thread barriers. This is some of the reasons why writing a high performance GPU kernel is often found to be a complex and time consuming task.

Because of the major overhead involved with writing GPU code we will, in the future, probably see a lot of work aimed at reducing it.  A recent development is the introduction of C++ AMP which enables developers to write accelerated code (e.g. by a GPU) in plain C++. Hopefully, in the future, GPU programming will be a job for compilers, and direct fiddling with GPU kernels should only be needed if maximum throughput is absolutely needed. This is similar to how SIMD instructions are handled in CPU-code today. Most of the time developers will rely on the CPU compilers to analyze the code and auto-apply SIMD instruction whenever it is possible. However, when maximum throughput is needed, developers with deep knowledge of the algorithm are often able to make faster code by manually writing SIMD instruction code. The relevance of auto-applied GPU instructions by compilers will also increase due to the recent embedding of GPUs on the same chip as the CPU. Soon, when writing code, we will probably talk about coding for a collection of heterogeneous cores where operations are automatically offloaded to the right core (e.g. a GPU) by the compiler.

@@

If the width of a piezoelectric element is less than the wavelength of the transmitted signal the element is said to be \textit{omnidirectional}, hence energy is distributed equally in all directions. When multiple omnidirectional elements are arranged in an array the transmitted pulse will get shaped into a beam\footnote{A beam refers to the pulse integrated in time.} of ultrasound by diffraction as the size (aperture) of the array grows. The 3 dB angular width of this beam, and also the angular resolution of the array, is given by \cite{AngelUltrasound}:
\begin{align}
\delta\theta = \frac{\lambda}{D},
\end{align}
\nomi{$\delta\theta$}{Angular resolution} where \nom{$\lambda$}{Wavelength} is the ultrasound wavelength, and \nom{$D$}{Aperture width} is the aperture width. With a speed-of-sound in human tissue of $1540\frac{\text{m}}{\text{s}}$ the angular resolution of a 2 cm cardiac array at 3 MHz is around 1.5\degree, which at 8 cm gives a lateral resolution of around 2 mm. The axial resolution is inverse proportional to the pulse length, and is typically better than the lateral resolution.

If time delays are applied on the signal sent to the different elements it is possible to steer and focus the ultrasound beam in space. This technique is know as \textit{beamforming}, and for ultrasound imaging it is applied both when transmitting and receiving ultrasound pulses. For medical ultrasound imaging there are two main types of ultrasound probes, linear and phased arrays \cite{AngelUltrasound}. A cardiac ultrasound array is a phased array, which means that the image is constructed from beams, or A-mode images, which are gradually steered from $-\theta/2$ to $\theta/2$, resulting in an imaging sector which are $\theta$ wide. This kind of image is known as a B-mode image (B for brightness). If \nom{$M$}{Number of array elements} is the number of elements in the array, and \nom{$\Delta_m[n]$}{Per-element focusing and steering delays} is the per-element focusing and steering delay, the standard form of beamforming used on receive, known as \textit{delay-and-sum}, is given as:
\begin{align}\label{eq:das}
z[n] = \sum_{m = 0}^{M-1}w_m^*x_m[n - \Delta_m[n]] = \vec{w}^H\vec{x}[n],
\end{align}
where \nom{$\vec{w}$}{Array weight vector (apodization)} is know as the array weight vector or apodization, and $z[n]$ is the output A-mode image.  If $p_o$ is the phase center of our array and $p$ is the point where we want to steer and focus the ultrasound beam, we have to use the following delays:
\begin{align}
\Delta_m[n] =\frac{|p_m+p| - |p_o+p|}{c},
\end{align}
where \nom{$c$}{Speed of sound} is the speed of sound in human tissue, and $p_m$ is the position of the m'th element. On transmit the inverse set of delays have to be used. Also note that on transmit we have to decide where to steer and focus prior to transmitting a pulse, transmit delays are therefore not time varying for an A-mode image.

>>>>

Modern GPUs can also be interpreted as a unit consisting of multiple SIMD cores (Fig.\,\ref{fig:cpu_gpu}). The recent Kepler architecture by Nvidia have up to 15 SIMD units capable of processing 192, 32 bit, floating point values per clock cycle at 745 MHz\footnote{Nvidia typically refers to a core as a unit capable of performing a FMA instruction on one single precision floating point value. Hence, in Nvidia's world, Kepler has $15*192=2880$ cores.}. These GPUs are also capable of performing FMA instructions, effectively doubling the number of theoretical FLOPS. The theoretical peak single precision FLOPS for this architecture is by that found to be 
\begin{equation}
15*192*2*745 = 4291\,\text{GFLOPS}.
\end{equation} 
Each SIMD core also comprise 16 special function units (\nom{SFU}{Special function unit (sin, exp, sqrt, ect.)}) which add close to 1 TFLOPS in additional computational power. This adds up to a total of five TFLOPS as shown in Fig.\,\ref{fig:cpu_vs_gpu} for the GeForce GTX 780 TI GPU. From Fig.\,\ref{fig:cpu_gpu} we see how the high number of SIMD cores found on GPUs and their larger width, compared with the SIMD cores on CPUs, results in five times the computational performance of a high-end CPU (without the integrated GPU).

@@

Both CUDA and OpenCL is centered around a kernel function which are executed by a large group of threads. The kernel function, for both CUDA and OpenCL, has a C-like syntax with some additional specifiers. An example of a CUDA kernel is depicted in Fig.\,\ref{fig:gpu_kernel}. The \texttt{\_\_global\_\_} specifier tells the CUDA compiler that this function is a kernel function, and in CUDA, this kernel function is launched by using a CUDA specific syntax:
\begin{align*}
\texttt{square<<<grid, block>>>(a, N)},
\end{align*}
where \texttt{grid} and \texttt{block} is 3-component vectors specifying grid and block size in "xyz" as depicted in Fig.\,\ref{fig:gpu_grid}, \texttt{a} is a pointer to a memory buffer located on the GPU, and \texttt{N}$*$\texttt{N} is the buffer size. When launched, a copy of the kernel is scheduled  for each thread in each block, and each kernel gets as input its location in the compute grid (\texttt{blockIdx} and \texttt{threadIdx}) together with the grid size (\texttt{gridDim} and \texttt{blockDim}). This information is then typically used to calculate which memory elements to gather from global memory, in this case the elements in \texttt{a}. %Behind the scene, thread blocks are divided into smaller sets of threads called a warp (or wavefront in OpenCL). 
When a kernel is executed, multiple thread blocks are scheduled to each SIMD core where they stay until all instructions in the kernel is finished for all threads in the block. Thus,  in order to formulate a given problem as a kernel function the global memory input to each thread must not rely on the output from other blocks, hence it must be possible for the GPU to scheduled blocks in any order.

As we see, a thread has a different conceptual meaning on a GPU than on a CPU.  Where a CPU is said to execute threads which can perform SIMD instructions, a thread on a GPU can be interpreted as stream capable of processing one data element in the SIMD instruction. This is why GPU threads are said to be light weighted. This conceptual difference also means that there is an inherent focus on breaking algorithms down and into SIMD instructions just by thinking of threads.

As the width of SIMD instructions increase the only algorithms that will benefit is those who can be divided into enough concurrent parts. Also, to achieve peak performance, the implementation must perform one FMA operation on each core per clock cycle and all memory latency must be hidden by instructions. This is something which is impossible to achieve in real life applications.

The code in Fig.\,\ref{fig:gpu_grid} is of course a minimal example of GPU programing. A full example will involve several lines of code transferring data to and from the GPU, and the kernel could include manual administration of the L1 cache (shared memory), and intra-block thread barriers. This is some of the reasons why writing a high performance GPU kernel is often found to be a complex and time consuming task. Because of the major overhead involved with writing GPU code we will, in the future, probably see a lot of work aimed at reducing it.  A recent development is the introduction of C++ AMP which enables developers to write accelerated code (e.g. by a GPU) in plain C++. Hopefully, in the future, GPU programming will be a job for compilers, and direct fiddling with GPU kernels should only be needed if maximum throughput is absolutely needed. This is similar to how SIMD instructions are handled in CPU-code today. Most of the time developers will rely on the CPU compilers to analyze the code and auto-apply SIMD instruction whenever it is possible. However, when maximum throughput is needed, developers with deep knowledge of the algorithm are often able to make faster code by manually writing SIMD instruction code. The relevance of auto-applied GPU instructions by compilers will also increase due to the recent embedding of GPUs on the same chip as the CPU. Soon, when writing code, we will probably talk about coding a heterogeneous collection of cores where operations are automatically offloaded to the right core (e.g. a GPU) by the compiler.

@@

If the width of a piezoelectric element is less than the wavelength of the transmitted signal the element is said to be \textit{omnidirectional}, hence energy is distributed equally in all directions. When multiple omnidirectional elements are arranged in an array the transmitted pulse will get shaped into a beam\footnote{A beam refers to the pulse in space integrated across time.} of ultrasound by diffraction as the size of the array  (the \textit{aperture}) grows. The 3 dB angular width of this beam, and also the angular resolution of the array, is given by \cite{AngelUltrasound}:
\begin{align}
\delta\theta = \frac{\lambda}{D},
\end{align}
\nomi{$\delta\theta$}{Angular resolution} where \nom{$\lambda$}{Wavelength} is the ultrasound wavelength, and \nom{$D$}{Aperture size} is the aperture width. With an average speed-of-sound in human tissue of $1540\frac{\text{m}}{\text{s}}$ the angular resolution or opening angle of a 2 cm cardiac probe operating at 3 MHz is around 1.5\degree, which at 8 cm gives a lateral resolution of around 2 mm. The axial resolution is inverse proportional to the pulse length, and is typically better than the lateral resolution.

If time delays are applied on the signal sent to the different elements it is possible to steer and focus the ultrasound beam in space. This technique is know as \textit{beamforming}, and for ultrasound imaging it is applied both when transmitting and receiving ultrasound pulses. For medical ultrasound imaging there are two main types of ultrasound probes, linear and phased probes \cite{AngelUltrasound}. A cardiac probe is a phased array, which means that the image is constructed from beams, or A-mode images, which are gradually steered from $-\theta/2$ to $\theta/2$ radians, resulting in an imaging sector which are $\theta$ radians wide. This kind of image, where A-mode images are acquired from multiple spatial locations in the xz-plane, is known as a B-mode image (B for brightness). If A-mode images are steered in the yz-plane as well we get what is known as a volumetric or 3D ultrasound image.

If \nom{$M$}{Number of array elements} is the number of elements in the array, and \nom{$\Delta_m[n]$}{Per-element focusing and steering delays} is the per-element focusing and steering delay, the standard form of beamforming used on receive, known as \textit{delay-and-sum}, is given as:
\begin{align}\label{eq:das}
z[n] = \sum_{m = 0}^{M-1}w_m^*x_m[n - \Delta_m[n]] = \vec{w}^H\vec{x}[n],
\end{align}
where \nom{$\vec{w}$}{Array weight vector (apodization)} is know as the array \textit{weight vector} or \textit{apodization}, and $z[n]$ is the output A-mode image in the direction dictated by $\Delta_m[n]$.  If $p_o$ is the phase center of our array and $p$ is the point where we want to steer and focus the ultrasound beam, we have to use the following delays:
\begin{align}
\Delta_m[n] =\frac{|p_m+p| - |p_o+p|}{c},
\end{align}
where \nom{$c$}{Speed of sound} is the speed of sound in human tissue, and $p_m$ is the position of the m'th element. On transmit the inverse set of delays have to be used. Also note that on transmit we have to decide where to steer and focus prior to transmitting a pulse, transmit delays are therefore not time varying for an A-mode image.
